\documentclass[12pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{\textbf{PHP Basic Syntax – Study Notes}}
\lhead{Author: \textbf{Mitegab}}
\rfoot{\thepage}

\title{\textbf{PHP Basic Syntax}}
\author{Mitegab}
\date{\today}

\begin{document}
\maketitle

\section*{Introduction}
PHP (Hypertext Preprocessor) is a widely used, open source scripting language especially suited for web development.  
This note summarizes the \textbf{basic syntax rules} and best practices for writing PHP code.

\section{PHP Tags}
\begin{itemize}[left=1em]
  \item PHP code is enclosed between \verb|<?php ... ?>|.
  \item Content outside these tags is sent directly to the browser (HTML or plain text).
  \item \textbf{Whitespace rule:} A space, tab, or newline should follow \verb|<?php| to ensure proper token separation.
  \item \textbf{Short echo tag:} \verb|<?= ... ?>| is shorthand for \verb|<?php echo ... ?>|.
  \item \textbf{Short open tags:} \verb|<? ... ?>| only work if \texttt{short\_open\_tag} is enabled in \texttt{php.ini} or compiled accordingly.  
        Recommended: stick to full tags \verb|<?php ... ?>| and \verb|<?= ... ?>| for portability.
\end{itemize}

\subsection*{Examples}
\begin{verbatim}
<?php echo 'Hello World'; ?>

<?= 'Short echo works too'; ?>

<? echo 'Short tags depend on php.ini'; ?>
\end{verbatim}

\section{Closing Tag Best Practice}
\begin{itemize}[left=1em]
  \item If a file ends with PHP code, omit the closing \verb|?>| tag to avoid accidental whitespace output.
  \item This is especially useful when using \texttt{include} or \texttt{require} and when sending HTTP headers later.
\end{itemize}

\subsection*{Example}
\begin{verbatim}
<?php
echo "Hello world\n";

// ... more code

echo "Last statement\n";
// Script ends with no closing tag
\end{verbatim}

\section{Embedding PHP in HTML (Escaping)}
\begin{itemize}[left=1em]
  \item Everything outside \verb|<?php ... ?>| is ignored by the PHP parser.
  \item This allows PHP to be embedded in HTML templates easily.
\end{itemize}

\subsection*{Example}
\begin{verbatim}
<p>This is ignored by PHP.</p>
<?php echo 'This is parsed.'; ?>
<p>This is also ignored by PHP.</p>
\end{verbatim}

\section{Advanced Escaping with Conditions}
\begin{itemize}[left=1em]
  \item You can drop out of PHP mode within conditional structures.
  \item PHP skips the HTML blocks where the condition is not met.
\end{itemize}

\subsection*{Example}
\begin{verbatim}
<?php if ($expression == true): ?>
  This will show if expression is true.
<?php else: ?>
  Otherwise this will show.
<?php endif; ?>
\end{verbatim}

\textbf{Tip:} For large text blocks, dropping out of PHP mode is more efficient than echoing everything.

\section{Instruction Separation (Semicolons)}
\begin{itemize}[left=1em]
  \item Like C or Perl, PHP statements must end with a semicolon.
  \item The closing \verb|?>| tag automatically implies a semicolon.
  \item The closing tag will also include the immediately trailing newline if present.
\end{itemize}

\subsection*{Examples}
\begin{verbatim}
<?php echo "Some text"; ?>
No newline
<?= "But newline now" ?>
\end{verbatim}

Outputs:
\begin{verbatim}
Some textNo newline
But newline now
\end{verbatim}

\section{Comments in PHP}
PHP supports C, C++ and shell-style comments:
\begin{itemize}[left=1em]
  \item \verb|// This is a one-line (C++) comment|
  \item \verb|# This is a one-line shell style comment|
  \item \verb|/* Multi-line comment */|
\end{itemize}

\subsection*{Example}
\begin{verbatim}
<?php
echo "This is a test\n"; // One-line C++ style
/* This is a multi-line comment
   yet another line */
echo "Another test\n";
echo "Final test\n"; # Shell-style comment
?>
\end{verbatim}

\textbf{Important:}
\begin{itemize}[left=1em]
  \item One-line comments end at the end of the line or at the next \verb|?>|.
  \item HTML after \verb|// ... ?>| or \verb|# ... ?>| will still be printed (because PHP exited).
  \item Do not nest \verb|/* */| comments; the first \verb|*/| ends the comment.
\end{itemize}

\subsection*{Problematic Example (Nested Comments)}
\begin{verbatim}
<?php
 /*
    echo 'This is a test'; /* This comment will cause a problem */
 */
?>
\end{verbatim}

\section{Additional Tips and Best Practices}
\begin{itemize}[left=1em]
  \item Use UTF-8 encoding without BOM to avoid hidden output before headers.
  \item Use \texttt{declare(strict\_types=1);} at the top of files if using PHP 7+ for strict typing.
  \item Always validate and escape user input (security).
  \item Follow PSR standards (PSR-1, PSR-12) for coding style consistency.
\end{itemize}

\section*{Summary}
\begin{itemize}[left=1em]
  \item Always use \verb|<?php ... ?>| and \verb|<?= ... ?>| for compatibility.
  \item Prefer omitting closing tag at end of PHP-only files.
  \item Drop out of PHP for large HTML blocks.
  \item End statements with semicolons.
  \item Use proper comment styles and avoid nesting block comments.
\end{itemize}

\bigskip
\noindent\textbf{Reference:} \href{https://www.php.net/manual/en/language.basic-syntax.php}{PHP Manual – Basic Syntax}

\end{document}
